Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model

### 2022-07-27 Rmarkdown/paper.Rmd
remove fig 9

<!-- to do: fix figure order -->

```{r setup, include=FALSE}
here::i_am("Rmarkdown/paper.Rmd")
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("stringr")
library("janitor")
model_in <- readr::read_csv(here::here("data-clean/model_in.csv"))
fig7_summ <- readr::read_csv(here::here("data-clean/figure-7-summary.csv"))
```

@@

Pipe loop studies are used to evaluate corrosion control treatment, and updated regulatory guidance will ensure that they remain important for water quality management. But the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common features that make popular methods, such as the *t*- or rank-sum tests, poor descriptive models. Here, we propose a model for pipe loop data that accommodates many of these difficult-to-model characteristics: a robust Bayesian generalized additive model with continuous-time autoregressive errors. Our approach facilitates corrosion control treatment comparisons without the need for imputing non-detects or special handling of outliers. It is well-suited to describing nonlinear trends without overfitting, and it accounts for the reduced information content in autocorrelated time series. We demonstrate it using a four-year pipe loop study, with multiple pipe configurations and orthophosphate dosing protocols, finding that an initially high dose of orthophosphate (2 mg P L^-1^) that is subsequently lowered (0.75 mg P L^-1^) can yield lower lead release than an intermediate dose (1 mg P L^-1^) in the long term.

@@

```{r figure-1, fig.cap="**Figure 1. (a)** An example time series from the study described in this paper; labels 1--5 represent features of the series that represent statistical modeling challenges. **(b)** Summary of the five components comprising the model fitted to pipe loop data."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-1.png"))
```

@@

Censoring typically arises in pipe loop studies when analyte concentrations fall below below the detection limit (i.e., left-censoring). Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values. Neither approach is advisable as both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012]. Popular statistical tests (e.g., the paired t-test and its generalization, the one-way repeated measures ANOVA) cannot accommodate censoring [@helsel_statistics_2012]. The non-parametric rank-sum test does allow for left-censoring at one detection limit, but it is not adequate for more complex censoring schemes (e.g., multiple detection limits or interval censoring). Imputation methods such as regression on order statistics [@helsel_statistics_2012] may be useful, but in a time series the missing data may be correlated, and imputation in the context of a full data model is usually preferable [@hopke_multiple_2001].

@@

Autocorrelation means the current state of the system depends on its history. For example, the effect of a perturbation to corrosion scale is evident for some time after the fact. From a statistical perspective, autocorrelation makes parameter estimates more uncertain [@helsel_statistical] by reducing the effective sample size (i.e., the information content of the sample) [@gelman_bayesian_2014]. It is very likely to be a feature of any pipe loop study dataset where effluent from pipes is sampled repeatedly over time, but it is not discussed in popular guides for pipe loop data analysis [@wysock_statistical_1995; @kirmeyer1996corr]. Moreover, neither the t-test nor its non-parametric analogues account for autocorrelation explicitly [@helsel_statistical; @raadt_pressing_2019].

Autocorrelation overstates the precision of statistical estimates [@gelman_bayesian_2014]. As a consequence, it inflates the probability of falsely rejecting null hypotheses [@helsel_statistical]. The effect of autocorrelation can be surprisingly large: 100 observations from a time series may provide no greater precision than 10 independent observations if the 100 are strongly correlated in time [@helsel_statistical]. That is, *n* autocorrelated observations provide less information than *n* independent observations, and the difference can be large if autocorrelation is strong [@gelman_bayesian_2014]. Accounting for autocorrelation can improve predictive accuracy, but if it is ignored, analysts run the risk of making incorrect decisions based on the data.

@@

Extreme lead or copper concentrations---typically due to particle release---are often inconsistent with the normality assumption common to many popular statistical tests. Outliers are sometimes omitted [@schock_causes_1990], but unless they represent measurement error or a separate population (e.g., a process error), it is usually preferable to use a statistical method that is robust against them [@helsel_statistical].

@@

Here, we propose a Bayesian approach for building statistical models of pipe loop study data that is sufficiently flexible for most pipe loop datasets. And while time is the only covariate in our model, additional predictors can easily be added as smooth or linear functions. The collection of measurement series representing variation in a given parameter over time---often lead--- is represented as the sum of five components (Figure 1b):

1. An average representing each pipe
2. A global trend describing multi-year variation in the parameter of interest across all pipes
3. A set of pipe-specific trends that describe the deviations from the global multi-year trend particular to each individual pipe
4. A seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it)
5. An autocorrelated error term

Separation of the time-trend into multi-year and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the multi-year trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are longer data gaps the series representing one or more pipes. In these instances, the model "borrows" information from the global term. The autocorrelated error model improves the accuracy of parameter distributions and helps to prevent overfitting.

We demonstrate the model using the data from a four-year pipe loop study that included multiple pipe configurations and orthophosphate dosing protocols. Our model estimates autocorrelation, despite an irregular sampling frequency; it also accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary.

We use the model to (1) compare corrosion control treatments in terms of dissolved and particulate lead, (2) compare their relative contributions to total lead release under each treatmenet, and (3) identify periods of stable lead release by considering the rate of change in the global multi-year term.

@@

## Summary

@@

Comparisons of corrosion control treatments are often aided by a statistical model. Typical choices in pipe loop studies include constant (*e.g.*, *t*-test) and linear models. But when applied to an example time series from our dataset (Figure 2a), both of these models violate a key assumption: that residuals are independent.

Lack of independence is apparent in Figure 2b: the residual of either model at time *t*, $\epsilon_t$, is correlated with the residual at time *t-1*, $\epsilon_{t-1}$. This can be extended to include the residual at time *t-k*, $\epsilon_{t-k}$, where *k = 1...21*, yielding the autocorrelation function (Figure 2c). Independent residuals yield autocorrelation functions that hover around zero, whereas correlated residuals typically yield autocorrelation functions that either decay or oscillate. We assume here that the example time series is a regular sequence, but in this case, that is only an approximation, and we relax it in the next section.

@@

The models included one-dimensional global smoothers describing multi-year variation in lead across all time series (Figure 4b). The model of dissolved lead also included a global smoother describing seasonal variation (Figure 4a). Both models included group-level smoothers describing multi-year variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed elsewhere [@pedersen_hierarchical_2019]. The multi-year smoothers were thin-plate regression splines and the seasonal smoother was a cyclic cubic regression spline [@mgcv]. To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance. To predict total lead, the particulate and dissolved predictions can be summed.

@@

Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F(x)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq x)$, the probability that, given a set of distributional parameters, a random variable $X$ is equal to the censoring limit, $x$, *at most*) [@stan2021; @mcelreath_statistical_2016]. Interval-censored observations are indicated by $censored = 2$ and were modeled using $G(x, y)$, or $P(x \lt X \leq y) = F(y) - F(x)$, where $x$ and $y$ are the lower and upper limits, respectively. Parameters, then, are estimated without imputation of nondetects: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(x)$, only.

@@

$$
(6)~~
\begin{align}
\alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha)~~[adaptive~prior] \\
\alpha \sim T(0.1, 2.5, 3)~~[prior~on~average~pipe] \\
\sigma_{\alpha} \sim T(0, 2.5, 3)~~[prior~on~\sigma~of~pipes] \\ 
\end{align}
$$
The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$; $r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$. The $f_j(t)$ are smooth spline functions that comprise the GAM, and each of the $n$ smoothers takes the following form:

@@

The posterior predictive distribution was also robust against extreme values (Figure 5d): the Pareto-*k* values shown here represent the influence of individual observations on the posterior distribution, and values below 0.7 are considered moderately influential, yielding reliable estimates of out-of-sample predictive accuracy (see *Comparing corrosion control treatments* below) [@loo; @loo_paper]. Here, all Pareto-*k* values were below 0.5.

@@

Posterior predictions from the models---$\mu_t$ in Equation $(5)$---can be compared to evaluate the orthophosphate dosing protocols. Here we compared the 0--0.5 mg P L^-1^ with the 1 mg P L^-1^ and the 1 mg P L^-1^ with the 2--0.75 mg P L^-1^ dosing protocols over time by considering their ratios.

The three dosing protocols yielded distinct lead release distributions over most of the study, and the difference between the lowest orthophosphate dose and the other two is particularly apparent (Figure 7). The ratio of dissolved Pb concentrations at a P ratio of 0:1---comparing the 0 mg P L^-1^ control with the 1 mg P L^-1^ dose---peaked at `r signif(fig7_summ$.epred, 3)` with a 95% credible interval of `r signif(fig7_summ$.lower, 3)`--`r signif(fig7_summ$.upper, 3)` (Figure 7a). The dissolved Pb ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^ (i.e., at a 0.5:1 P ratio), but its 2.5th percentile---the lower bound of the 95% credible interval---remained above 1 with a single brief exception. This highlights the superior corrosion control performance of 1 mg P L^-1^ over both 0 and 0.5 mg P L^-1^ and is consistent with expectations based on geochemical solubility models [@schock1996corr].

Pb ratios comparing the 1 and 2--0.75 mg P L^-1^ protocols exhibited a more nuanced pattern (Figure 7b). As a comparison of either particulate or dissolved lead concentrations, the Pb ratio at a 1:2 P ratio was less than one, suggesting that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose. This likely occurred in the dissolved fraction as small colloids [@trueman_characterizing_2019; @li_impact_2020; @lytle_lead_2020].

The transition from 2--0.75 mg P L^-1^ reversed this pattern: the dissolved and particulate Pb ratios at the 1:0.75 P ratio were largely greater than one. A high dose of 2 mg P L^-1^, then, presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^, despite being exposed to a lower concentration of 0.75 mg P L^-1^. This implies a potential advantage and drawback of a high passivation dose, followed by a lower maintenance dose, of orthophosphate. That is, a passivation dose may cause excess lead release in the short term---likely due to particle-generating mechanisms [@lytle_lead_2020; @zhao_formation_2018]---while ensuring that a lower maintenance dose is either more effective or effective sooner than it otherwise might be.

@@

Autocorrelation models the dependence of observations on previous observations, or the response of the series at time $t+s$ to a shock at time $t$. The effect of the autocorrelated error term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov. 2020 ("Pb-Cu", Figure 6). While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4). That is, extreme and sustained deviations from typical behavior are generally captured by the error term (but they are still reflected in model predictions: see Figure 6 and equation 5).

Accounting for autocorrelation also improved the predictive accuracy of the models, assessed here using approximate leave-one-out cross-validation on the complete dataset. Approximate leave-one-out cross-validation estimates the expected log pointwise predictive density---a metric of model accuracy---for new observations not used to fit the model [@loo; @loo_paper; @mcelreath_statistical_2016]. We applied it to the particulate and dissolved lead models with continuous-time autoregressive errors and also to identically-specified models with independent errors. The autoregressive error models yielded higher expected log pointwise predictive densities (i.e., greater accuracy), exceeding the independent error models in this metric by 206 and 47 (dissolved and particulate lead models, respectively).

## Estimating the particulate fraction

Across all pipe sections, a majority of total lead was present in 0.45 µm filtrate (Figure 9). Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible. During the conditioning period---before orthophosphate was introduced---particle release depended mainly on the pipe configuration, with galvanically-coupled lead and copper ("Pb-Cu") yielding the greatest particulate fraction. Almost immediately after conditioning, the particulate fraction increased in the first two arrays as orthophosphate decreased lead solubility. The inverse seasonality---higher particulate fractions in winter---is a function of the relatively constant particulate lead release and the strongly seasonal dissolved lead release (Figure 6). The particulate fraction was also higher in the effluent from pipes featuring a galvanic junction. This, and the effect of orthophosphate on lead speciation, is consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020].

```{r figure-9, fig.cap="**Figure 9.** The particulate fraction of total lead over time, grouped by pipe and orthophosphate dosing protocol, and estimated from the posterior distributions of the models fitted to particulate and dissolved lead time series. The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-9.png"))
```

@@

The variation in dissolved lead captured in Figure 8a neglects seasonality by design. And while the seasonal component of the model (Figure 8b) is described as a function of ordinal day, this is a proxy for the physical and chemical inputs that are expected to have caused it. The true seasonal relationship is probably complex and difficult to determine from the study data, but temperature and influent iron concentration are plausible drivers: both varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016]. And while dissolved organic carbon has also been shown to impact lead release [@korshin_influence_2005; @lin_release_2008; @king_role_2022], true colour---a proxy for dissolved organic carbon [@redden_chemical_2021]---was not strongly seasonal (Figure S9).

@@

The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the large difference in lead release between the 0 and 1 mg P L^-1^ dose. The effect of orthophosphate dose was complex: lead release at 1 mg P L^-1^ was lower than at 0.5 mg P L^-1^ but higher than at 0.75 mg P L^-1^. We attribute the superior performance of the 0.75 mg P L^-1^ dose to an initial high dose of 2 mg P L^-1^. On the other hand, the initial 2 mg P L^-1^ dose caused excess lead release, highlighting an important trade-off that should be considered in applying a passivation-maintenance (high-low) dosing strategy. And since high orthophosphate doses may generate small lead-orthophosphate colloids/nanoparticles that are difficult to remove from drinking water [@lytle_lead_2020], the 1 mg P L^-1^ dose is the conservative choice. The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring. This seasonal pattern may have multiple drivers: iron and water temperature, for instance, were also seasonal.

>>>>

<!-- to do: check figure order -->
<!-- remove figure 9 from figures.R-->

```{r setup, include=FALSE}
here::i_am("Rmarkdown/paper.Rmd")
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("tidyr")
library("stringr")
library("janitor")
model_in <- readr::read_csv(here::here("data-clean/model_in.csv"))
fig7_summ <- readr::read_csv(here::here("data-clean/figure-7-summary.csv"))
```

@@

Pipe loop studies are used to evaluate corrosion control treatment, and updated regulatory guidance will ensure that they remain important for water quality management. But the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common attributes that make popular methods, such as the *t*- or rank-sum tests, poor descriptive models. Here, we propose a model for pipe loop data that accommodates many of these difficult-to-model attributes: a robust Bayesian generalized additive model with continuous-time autoregressive errors. Our approach facilitates corrosion control treatment comparisons without the need for imputing non-detects or special handling of outliers. It is well-suited to describing nonlinear trends without overfitting, and it accounts for the reduced information content in autocorrelated time series. We demonstrate it using a four-year pipe loop study, with multiple pipe configurations and orthophosphate dosing protocols, finding that an initially high dose of orthophosphate (2 mg P L^-1^) that is subsequently lowered (0.75 mg P L^-1^) can yield lower lead release than an intermediate dose (1 mg P L^-1^) in the long term.

@@

@@

Censoring typically arises in pipe loop studies when analyte concentrations fall below the detection limit (i.e., left-censoring). Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values. Neither approach is advisable as both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012]. Popular statistical tests (e.g., the paired t-test and its generalization, the one-way repeated measures ANOVA) cannot accommodate censoring [@helsel_statistics_2012]. The non-parametric rank-sum test does allow for left-censoring at one detection limit, but it is not adequate for more complex censoring schemes (e.g., multiple detection limits or interval censoring). Imputation methods such as regression on order statistics may be useful [@helsel_statistics_2012], but in a time series the censored data may be correlated, and imputation in the context of a full data model is usually preferable [@hopke_multiple_2001].

@@

Autocorrelation means the current state of the system depends on its history. For example, the effect of a perturbation to corrosion scale is evident for some time after the fact. From a statistical perspective, autocorrelation makes parameter estimates more uncertain by reducing the effective sample size (i.e., the information content of the sample) [@helsel_statistical; @gelman_bayesian_2014]. As a consequence, it inflates the probability of falsely rejecting null hypotheses [@helsel_statistical]. Autocorrelation is very likely to be a feature of any pipe loop study dataset where effluent from pipes is sampled repeatedly over time, but it is not discussed in popular guides for pipe loop data analysis [@wysock_statistical_1995; @kirmeyer1996corr]. Moreover, neither the t-test nor its non-parametric analogues account for autocorrelation explicitly [@helsel_statistical; @raadt_pressing_2019].

The effect of autocorrelation can be surprisingly large: 100 observations from a time series may provide no greater precision than 10 independent observations if the 100 are strongly correlated in time [@helsel_statistical]. That is, *n* autocorrelated observations provide less information than *n* independent observations, and the difference can be large if autocorrelation is strong [@gelman_bayesian_2014]. Accounting for autocorrelation can improve predictive accuracy, but if it is ignored, analysts run the risk of making incorrect decisions based on the data.

@@

Extreme lead or copper concentrations---typically due to particle release---are often inconsistent with the normality assumption common to many popular statistical tests. Outliers are sometimes omitted in practice [@schock_causes_1990], but unless they represent measurement error or a separate population (e.g., a process error), it is usually preferable to use a statistical method that is robust against them [@helsel_statistical].

```{r figure-1, fig.cap="**Figure 1. (a)** An example time series from the study described in this paper; labels 1--5 denote attributes of the series presenting statistical modeling challenges. **(b)** Summary of the five components comprising the model fitted to pipe loop data."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-1.png"))
```

@@

Here, we propose a Bayesian generalized additive model for pipe loop study data that is sufficiently flexible for most pipe loop datasets. The collection of measurement series representing variation in a given parameter over time---often lead--- is represented as the sum of five components (Figure 1b):

1. An average representing each pipe.
2. A smooth seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (this component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it).
3. A smooth global trend describing multi-year variation in the parameter of interest across all pipes.
4. A set of smooth pipe-specific trends that describe the deviations from the global multi-year trend particular to each pipe.
5. An autocorrelated error term.

Separation of the time-trend into multi-year and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the multi-year trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are long data gaps in the series representing one or more pipes. In these instances, the model "borrows" information from the global term. The autocorrelated error model improves the accuracy of parameter distributions and predictions, while helping to prevent overfitting [@simpson_modelling_2018]. And while time is the only covariate in our model, additional predictors can easily be added as smooth or linear functions.

We demonstrate the model using the data from a four-year pipe loop study that included multiple pipe configurations and orthophosphate dosing protocols. Our model accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary.

We use the model to (1) compare corrosion control treatments in terms of dissolved and particulate lead, (2) compare their relative contributions to total lead release, and (3) identify periods of stable lead release by considering the rate of change in the global multi-year term.

@@

@@

Comparisons of corrosion control treatments are often aided by a statistical model, and typical choices in pipe loop studies include constant (*e.g.*, *t*-test) and linear models. But when applied to an example time series from our dataset (Figure 2a), both of these models violate a key assumption: the independence of model residuals.

Lack of independence is apparent in Figure 2b: the residual of either model at time *t*, $\epsilon_t$, is correlated with the residual at time *t-1*, $\epsilon_{t-1}$. This can be extended to include the residual at time *t-k*, $\epsilon_{t-k}$, where *k = 1...21*, yielding the autocorrelation function (Figure 2c). Independent residuals yield autocorrelation functions that remain close to zero, whereas correlated residuals yield autocorrelation functions exhibiting a pattern: decay or oscillation, for instance. We assume here that the example time series is a regular sequence, but in this case, that is only an approximation, and we relax it in the next section.

@@

The models included one-dimensional global smoothers describing multi-year variation in lead across all time series (Figure 4b). The model of dissolved lead also included a global smoother describing seasonal variation (Figure 4a). Both models included group-level smoothers describing multi-year variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed elsewhere [@pedersen_hierarchical_2019]. The multi-year smoothers were thin-plate regression splines and the seasonal smoother was a cyclic cubic regression spline [@mgcv]. To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance. While omitted here, total lead could be predicted using the same model structure.

@@

Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F(x)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq x)$, the probability that, given a set of distributional parameters, a random variable $X$ is equal to the censoring limit, $x$, *at most*) [@stan2021; @mcelreath_statistical_2016]. Interval-censored observations are indicated by $censored = 2$ and were modeled using $G(x, y)$, or $P(x \lt X \leq y) = F(y) - F(x)$, where $x$ and $y$ are the lower and upper limits, respectively. Parameters, then, were estimated without imputation of nondetects: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(x)$, only.

@@

$$
(6)~~
\begin{align}
\alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha)~~[adaptive~prior] \\
\alpha \sim T(0.1, 2.5, 3)~~[prior~on~average~pipe] \\
\sigma_{\alpha} \sim T(0, 2.5, 3)~~[prior~on~\sigma~of~pipes] \\ 
\end{align}
$$
The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$; $r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$. The $f_j(t)$ are smooth spline functions of time that comprise the GAM, and each of the $n$ smoothers takes the following form:

@@

The posterior predictive distribution was also robust against extreme values (Figure 5d): the Pareto-*k* values shown here represent the influence of individual observations on the posterior distribution, and values below 0.7 are considered moderately influential (at most), yielding reliable estimates of out-of-sample predictive accuracy (see *Comparing corrosion control treatments* below) [@loo; @loo_paper]. Here, all Pareto-*k* values were below 0.5.

@@

Posterior predictions from the models---$\mu_t$ in Equation $(5)$---can be compared to evaluate the orthophosphate dosing protocols. Here we compared the 0--0.5 mg P L^-1^ with the 1 mg P L^-1^ and the 1 mg P L^-1^ with the 2--0.75 mg P L^-1^ dosing protocols over time by considering their ratios. This highlights an advantage of the Bayesian approach---straightforward error propagation for quantities derived from the posterior distribution of the model parameters [@korner_nievergelt_bayesian_2015].

The three dosing protocols yielded distinct lead release distributions over most of the study, and the difference between the lowest orthophosphate dose and the other two is particularly apparent (Figure 7). The ratio of dissolved Pb concentrations at a P ratio of 0:1---comparing the 0 mg P L^-1^ control with the 1 mg P L^-1^ dose---peaked at `r signif(fig7_summ$.epred, 3)` with a 95% credible interval of `r signif(fig7_summ$.lower, 3)`--`r signif(fig7_summ$.upper, 3)` (Figure 7a). That is, lead release at 0 mg P L^-1^ was as much as 31 times higher than lead release at 1 mg P L^-1^.

The dissolved Pb ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^ (i.e., at a 0.5:1 P ratio), but its 2.5th percentile---the lower bound of the 95% credible interval---remained above 1 with a single brief exception. This highlights the superior corrosion control performance of 1 mg P L^-1^ over both 0 and 0.5 mg P L^-1^ and is consistent with expectations based on geochemical solubility models [@schock1996corr].

Pb ratios comparing the 1 and 2--0.75 mg P L^-1^ protocols exhibited a more nuanced pattern (Figure 7b). As a comparison of either particulate or dissolved lead concentrations, the Pb ratio at a 1:2 P ratio was less than one, meaning that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose. This likely occurred in the dissolved fraction as small colloids [@trueman_characterizing_2019; @li_impact_2020; @lytle_lead_2020].

The transition from 2--0.75 mg P L^-1^ reversed this pattern: the dissolved and particulate Pb ratios at the 1:0.75 P ratio were largely greater than one, meaning greater lead release at the 1 mg P L^-1^ dose. A high dose of 2 mg P L^-1^, then, presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^. This is despite being exposed to a lower orthophosphate concentration in the latter half of the study (0.75 mg P L^-1^), and it highlights a potential advantage and drawback of the passivation-maintenance dosing strategy. That is, a high passivation dose may cause excess lead release in the short term---likely due to particle-generating mechanisms [@lytle_lead_2020; @zhao_formation_2018]---while ensuring that a lower maintenance dose is either more effective or effective sooner than it otherwise might be.

@@

Autocorrelation models the dependence of observations on previous observations, or the response of the series at time $t+s$ to a shock at time $t$. The effect of the autocorrelated error term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov. 2020 ("Pb-Cu", Figure 6). While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4). That is, extreme and sustained deviations from typical behavior are generally captured by the error term (but still reflected in model predictions: see Figure 6 and equation 5).

Accounting for autocorrelation improved the predictive accuracy of the models, assessed here using approximate leave-one-out cross-validation on the full dataset. Approximate leave-one-out cross-validation estimates the expected log pointwise predictive density---a metric of model accuracy---for new observations not used to fit the model [@loo; @loo_paper; @mcelreath_statistical_2016]. We applied it to the particulate and dissolved lead models with continuous-time autoregressive errors and also to identically-specified models with independent errors. The autoregressive error models yielded higher expected log pointwise predictive densities (i.e., greater accuracy), exceeding the independent error models in this metric by 206 and 47 (dissolved and particulate lead models, respectively).

## Trends in particulate lead release

Over much of the study, the dissolved fraction (<0.45 µm) dominated total lead release (Figure 6). During the conditioning period in particular---before orthophosphate was introduced---particles represented a minor contribution. Orthophosphate treatment caused a sharp decrease in dissolved lead (<0.45 µm) and a concomitant increase in the particulate fraction, which is consistent with previous observations and geochemical models [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020]. Given the prominent seasonality in dissolved lead release, the particulate fraction exhibited inverse seasonality, with higher particulate fractions in winter (Figure 6). The increase in particulate lead in the final weeks of the study period is not easily explained, but it may be related to increasing feedwater aluminum concentrations (Figure S9)---previous work has shown that aluminum and orthophosphate interact to increase particulate lead release [@trueman_seasonal_2022].

@@

The variation in dissolved lead captured in Figure 8a neglects seasonality by design. And while the seasonal component of the model (Figure 8b) is described as a function of time (ordinal day), this is a proxy for the physical and chemical inputs that are expected to have caused it. The true seasonal relationship is probably complex and difficult to determine from the study data, but temperature and influent iron concentration are plausible drivers: both varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016]. And while dissolved organic carbon has also been shown to impact lead release [@korshin_influence_2005; @lin_release_2008; @king_role_2022], true colour---a proxy for dissolved organic carbon [@redden_chemical_2021]---was not strongly seasonal (Figure S9).

@@

The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the large difference in lead release between the 0 and 1 mg P L^-1^ dose. The effect of orthophosphate dose was complex: lead release at 1 mg P L^-1^ was lower than at 0.5 mg P L^-1^ but higher than at 0.75 mg P L^-1^. We attribute the superior performance of the 0.75 mg P L^-1^ dose to an initial high dose of 2 mg P L^-1^. On the other hand, the initial 2 mg P L^-1^ dose caused excess lead release, highlighting an important trade-off that should be considered in applying a passivation-maintenance (high-low) dosing strategy. And since high orthophosphate doses may generate small lead-orthophosphate colloids/nanoparticles that are difficult to remove from drinking water [@lytle_lead_2020], the 1 mg P L^-1^ dose may be a better option in this case. The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring. This seasonal pattern may have multiple drivers: iron and water temperature, for instance, were also seasonal.
